---
title: "Template Post"
author: "Frida Gomam"
date: 2015-07-23
tags: ["R Markdown", "plot", "regression"]
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

Each blog post is an R Markdown document. For more details on using R Markdown
see <http://rmarkdown.rstudio.com>. The first paragraph or so will appear on the
front page; by default the first 70 words are used, but you should instead write
the first paragraph so it can serve as a summary, and mark the split point
yourself like this:

<!--more-->

Notice that when we explicitly mark the split point, it does *not* appear in the
post itself, just on the front page. You should treat the paragraph above as an
abstract or teaser for the post, not as part of the post itself. Usually we do
not include links in the teaser.

You can embed an R code chunk like this:

```{r cars}
summary(cars)
fit <- lm(dist ~ speed, data = cars)
fit
```

## Including Plots

You can also embed plots. See Figure \@ref(fig:pie) for example:

```{r pie, fig.cap='A fancy pie chart.', tidy=FALSE}
par(mar = c(0, 1, 0, 1))
pie(
  c(280, 60, 20),
  c('Sky', 'Sunny side of pyramid', 'Shady side of pyramid'),
  col = c('#0292D8', '#F7EA39', '#C4B632'),
  init.angle = -50, border = NA
)
```

For posts where we're using our public data and showing it off, including some
of the code chunks would be great so that readers see how easy it is to use our
data. But for other audiences, it's probably best to use the `echo=FALSE` chunk
option so the code is not included in the post.

## Including Math

You can embed mathematics by using dollar signs for inline math and double
dollars for display math: $x = \sum_{i=1}^n i$,

$$
\frac{-b \pm \sqrt{b^2 - 4ac}}{2a}
$$

## Metadata

Each post has a title, author, date, and tags. The `draft` attribute marks a
post that should not be included in the rendered and published site, such as
this one. Choose bold, active titles, like "Delphi releases new survey data",
rather than boring titles like "New survey aggregates".

The author metadata should credit the post author or authors; as shown at the
end of this example post, you should also include a block about the authors and
linking to their home pages. But in the post text, you should also generously
name anyone who helped with parts of your post, e.g., the team members who
obtained data or set up a server or developed a package you use.

### Tags

Each post can be tagged, as you can see in the metadata block at the top. I
suggest we consider the following tags as base tags:

* forecasting
* nowcasting
* symptom surveys
* medical records
* COVIDcast API
* COVIDcast map
* data sources (for everthing else than symptom surveys and medical records)
* news (for announcements of new features, new models, etc.)
* R (for posts containing R, typically, our covidcast R package)
* Python (for posts containing Python, typically, our covicast Python package)

---

*[Frida Gomam](https://example.com) is a member of the Delphi group and a Ph.D.
student in data wrangling at the University of Southern North Dakota at Hoople.*
