---
title: "New COVID Symptom Surveys Track Testing, Mask-Wearing, and Mental Health"
author: "Alex Reinhart"
date: 2020-10-05
tags: ["symptom surveys"]
---

Beginning on September 8, 2020, we deployed a new version of our symptom survey.
Facebook helps us recruit tens of thousands of respondents daily, and the new
survey gives us unprecedented insights into the effects of COVID-19 across the
United States. Today we release new public datasets and share maps revealing
access to COVID testing, changes in mental health, and public use of masks.

<!--more-->

Beginning in early April 2020, the [Delphi group](https://delphi.cmu.edu/) has
conducted a major survey to track COVID-19 across the United States. With the
support of Facebook, we have been able to recruit tens of thousands of active
Facebook users to take our voluntary survey *every day*; meanwhile, a University
of Maryland team has conducted a [parallel international
effort](https://covidmap.umd.edu/) covering over 100 countries worldwide. Every
day, we aggregate our survey results to produce estimates of symptoms for
counties and states across the United States, making these estimates available
through our [COVIDcast map](https://covidcast.cmu.edu/) and our [public
API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast.html).

In previous posts, we described how [self-reported symptoms on our survey can
correlate strongly with COVID
cases](https://delphi.cmu.edu/blog/2020/08/26/covid-19-symptom-surveys-through-facebook/)
and how [symptom survey data could improve forecasts of COVID
cases](https://delphi.cmu.edu/blog/2020/09/21/can-symptoms-surveys-improve-covid-19-forecasts/).
We're also part of the [COVID-19 Symptom Data
Challenge](https://www.symptomchallenge.org/), which asked participants to use our
symptom survey data to enable earlier detection of outbreaks and help public
health authorities and the general public respond to the pandemic. (The first
submissions were due on October 6th, and we look forward to sharing the results.)

Beginning this summer, we began the process of revising our survey to address
new topics that had arisen as key problems for public health:

- What groups and occupations have been most affected by COVID-19?

- Who is being tested for COVID-19, and what are their test results?

- Do people who want COVID-19 tests have access to tests?

- Are people with symptoms seeking medical care? If so, what kind of care?

- Is the public following public health recommendations, such as the
  [recommendation to wear masks in
  public](https://www.cdc.gov/coronavirus/2019-ncov/prevent-getting-sick/diy-cloth-face-coverings.html)
  to reduce the spread of COVID-19?

- Are people participating in activities that may spread COVID-19, such as
  attending public events, working indoors with others, or going to school in
  person?

Some of these topics were partially addressed by our existing survey, but we
knew that as the pandemic changes and public health priorities adapt, our survey
must change also. We began a consultation process: In discussions with
researchers using our data, with public health officials, and with team members
at Carnegie Mellon and at Facebook, we established the most important topics to
address in the new survey, and began to draft questions. Through several rounds
of revisions, we finalized a new version of the survey that includes new
questions on testing, mask use, activities, and demographics.

(Meanwhile, questions about the public's beliefs, behaviors, and norms
surrounding COVID-19 are being addressed by a [separate survey conducted by
researchers at MIT](https://covidsurvey.mit.edu/), also in partnership with
Facebook to recruit respondents.)

## Insights from the New Survey Instrument

The full text of the survey is [available on our documentation
site](https://cmu-delphi.github.io/delphi-epidata/symptom-survey/coding.html)---this
version is Wave 4.

```{r sample-size, message=FALSE, include=FALSE}
library(covidcast)

start_day <- "2020-09-08"
end_day <- "2020-10-01"

data <- covidcast_signal("fb-survey", "raw_cli", start_day, end_day,
                         geo_type = "state")
sample_size <- prettyNum(signif(sum(data$sample_size), 3), big.mark = ",")
```

Between the new survey's deployment on September 8, 2020, and October 1st, we
have collected `r sample_size` valid responses from respondents across the
United States. Each response includes questions about symptoms, mask wearing,
testing, and the other important topics described above, along with demographic
details about the respondent. These demographics include age, gender, race,
occupation, and education, allowing us to understand how different groups have
been affected and which groups are currently most vulnerable to COVID-19. (As we
[described
before](https://delphi.cmu.edu/blog/2020/08/26/covid-19-symptom-surveys-through-facebook/),
Facebook does not receive any individual survey responses; we at Carnegie Mellon
collect this data and prepare aggregate data, which does not identify any
individual respondent, for public release.)

Let's explore just a few of the questions on the survey to see what they can
tell us. All of the plots below were made using our [covidcast R
package](https://cmu-delphi.github.io/covidcast/covidcastR/) using data we make
publicly available in the [COVIDcast
API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast.html)---click the
Code button to see the full code for each example.

### Mask Wearing

First, a simple question: What percentage of respondents say that they wear a
mask most or all of the time when they're in public?

```{r mask-wearing, message=FALSE}
library(covidcast)
library(ggplot2)

plot_label <- labs(caption = "Data from Delphi COVIDcast, covidcast.cmu.edu")
date_label <- labs(subtitle = format(as.Date(end_day), "%B %e, %Y"))

mask <- covidcast_signal("fb-survey", "smoothed_wwearing_mask",
                         start_day = end_day, end_day = end_day,
                         geo_type = "state")
plot(mask, range = c(55, 100),
     title = "% wearing masks in public most or all the time") +
    plot_label + date_label
```

Because the survey runs continuously, collecting data on mask-wearing every day,
we will be able to track how the prevalence of mask-wearing changes over time.
This may help public health officials target public awareness campaigns, and
also aid researchers trying to understand where the virus is most likely to
spread and what measures are most effective in preventing it.

### Testing

Access to COVID testing is important for many reasons: readily available tests
make it possible for cases to be detected sooner, before the virus can spread,
and help contact tracers do their work of tracking down new cases. Shortages of
tests can impede contact tracing and can also be harmful in settings like
nursing homes, where many residents are particularly vulnerable to COVID-19 and
staff must be tested regularly to ensure residents aren't exposed.

Testing data is published by state health authorities and aggregated by
organizations such as the [COVID Tracking Project](https://covidtracking.com/);
our own [COVIDcast map](https://covidcast.cmu.edu/) also includes data from
antigen tests produced by Quidel, a major test product manufacturer. But a
survey allows us to ask people directly whether they have been tested, and
allows us to understand *who* has been tested.

For example, we can ask: What percentage of people in each state say they were
tested in the last 14 days?

```{r pct-tested, message=FALSE}
tested <- covidcast_signal("fb-survey", "smoothed_wtested_14d",
                           start_day = end_day, end_day = end_day,
                           geo_type = "state")
plot(tested, range = c(6, 25),
     title = "% tested in the last 14 days") +
    plot_label + date_label
```

We can see a strong variation across the country---in North Dakota, roughly a
quarter of people were tested in the last 14 days!

It's worth thinking about this result in more detail. If we take North Dakota's
adult population, 25% of that population is roughly twice as many tests as North
Dakota's publicly reported testing data says were conducted during that time
period. But North Dakota's published testing figures [only include PCR tests,
not COVID antigen
tests](https://khn.org/news/lack-of-antigen-test-reporting-leaves-country-blind-to-the-pandemic/),
meaning the public figures leave out an important part of testing that the
survey data can illuminate. Antigen tests, and new rapid tests, will only become
more important in the coming months, making it important for epidemiologists to
have information about their use.

But testing volume is only one part of the issue. Another question is how many
people are testing positive. If a large percentage of tests have positive
results, that suggests that tests are only being given to people at high risk of
illness, rather than being used in mass testing campaigns. When survey
respondents say they were tested in the past 14 days, we also ask if they tested
positive:

```{r test-positivity, message=FALSE}
positivity <- covidcast_signal("fb-survey", "smoothed_wtested_positive_14d",
                               start_day = end_day, end_day = end_day,
                               geo_type = "state")
plot(positivity, range = c(4, 22),
     title = "% of tests that were positive, last 14 days") +
    plot_label + date_label
```

We can see quite a large variation in test positivity across the United States,
suggesting different access to testing.

But test positivity only indirectly answers a crucial question: If people *want*
a test, can they get it? When our survey respondents say they have *not* been
tested in the past 14 days, we ask whether they *wanted* to be tested in that
time, then ask why they were not. This also varies across the United States:

```{r wanted-test, message=FALSE}
wanted_test <- covidcast_signal("fb-survey", "smoothed_wwanted_test_14d",
                                start_day = end_day, end_day = end_day,
                                geo_type = "state")
plot(wanted_test, range = c(5, 12),
     title = "% not tested who wanted a test, last 14 days") +
    plot_label + date_label
```

In short, while the survey data does not replace official public health
reporting of COVID testing and case counts, it can provide insights not
available any other way. By providing these signals to the public in the
COVIDcast API, alongside numerous other data streams, we hope to provide
researchers, public health officials, and journalists the information they need
to form a more complete picture of the pandemic.


## You Can Help Analyze This Data

All of the maps and graphs above were built using data we make publicly
available through our COVIDcast API, and the exact technical details are
described in our [signal
documentation](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/fb-survey.html).
Many of these data streams will soon be featured on our [COVIDcast interactive
map](https://covidcast.cmu.edu/) as well.

But we also know that many researchers will have questions that can't be
answered from these simple county- and state-level averages. What occupations
have the highest rate of COVID-19? What age groups engage in the most behavior
at risk of transmitting COVID-19? What symptoms are most common in people who
report testing positive for COVID-19? Many of these questions require full
access to the survey data, including demographics about each respondent, to
answer. That is why we share de-identified survey data with academic and
nonprofit researchers studying the pandemic. We encourage all researchers to
[submit a simple
form](https://dataforgood.fb.com/docs/covid-19-symptom-survey-request-for-data-access/)
to get access.

Of course, we require all those with access to protect the confidentiality of
survey respondents, and permit the public release of only aggregate data that
can't be used to identify any respondent. And while Facebook supports this
project by helping recruit participants in the survey, they never receive
individual respondent's answers to survey questions, only aggregate data that
does not identify any individual respondents.

Between the data we make available through COVIDcast, and the many researchers
using our raw data to answer important questions, we hope this survey can serve
important roles informing our national pandemic response. Armed with the right
data, we can make decisions needed to protect public health and permit safe
reopening.

**Credits.** *Many items in Wave 4 of our survey are based on work by the team
at the Joint Program in Survey Methodology at the University of Maryland:
Adrianne Bradford and Samantha Chiu helped design items and gave feedback, led
by Frauke Kreuter. Facebook's survey team, including Sarah LaRocca and Katherine
Morris, gave extensive feedback on Wave 4 of the survey and assisted in its
deployment. Taylor Arnold, Alex Reinhart, and Kathryn Mazaitis developed the
code needed to process Wave 4.*

---

*[Alex Reinhart](https://www.refsmmat.com) manages Delphi's surveys, and is an
Assistant Teaching Professor in the Department of Statistics & Data Science at
CMU.*
