---
title: "COVIDcast Data API"
author: "Alex Reinhart and Katie Mazaitis"
date: 2015-09-07
tags: ["api", "api client"]
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

A teaser paragraph about the API

<!--more-->

* Our open source philosophy
* The purpose of the API
* Our unique data sources
* The data schema
* The clients and their features
* The backend
* A short note on Epidata history
* Conclude somehow

---


*[Alex Reinhart](https://www.refsmmat.com) manages Delphi's surveys, and is an
Assistant Teaching Professor in the Department of Statistics & Data Science at
CMU.*

*[Kathryn Mazaitis](https://cs.cmu.edu/~krivard) leads Delphi's
engineering team, and is a Senior Research Programmer in the Machine
Learning Department at CMU.*

---

Remaining template content for instructive purposes:

You can embed an R code chunk like this:

```{r cars}
summary(cars)
fit <- lm(dist ~ speed, data = cars)
fit
```

## Including Plots

You can also embed plots. See Figure \@ref(fig:pie) for example:

```{r pie, fig.cap='A fancy pie chart.', tidy=FALSE}
par(mar = c(0, 1, 0, 1))
pie(
  c(280, 60, 20),
  c('Sky', 'Sunny side of pyramid', 'Shady side of pyramid'),
  col = c('#0292D8', '#F7EA39', '#C4B632'),
  init.angle = -50, border = NA
)
```

For posts where we're using our public data and showing it off, including some
of the code chunks would be great so that readers see how easy it is to use our
data. But for other audiences, it's probably best to use the `echo=FALSE` chunk
option so the code is not included in the post.

## Including Math

You can embed mathematics by using dollar signs for inline math and double
dollars for display math: $x = \sum_{i=1}^n i$,

$$
\frac{-b \pm \sqrt{b^2 - 4ac}}{2a}
$$

## Metadata

Each post has a title, author, date, and tags. The `draft` attribute marks a
post that should not be included in the rendered and published site, such as
this one. Choose bold, active titles, like "Delphi releases new survey data",
rather than boring titles like "New survey aggregates".

The author metadata should credit the post author or authors; as shown at the
end of this example post, you should also include a block about the authors and
linking to their home pages. But in the post text, you should also generously
name anyone who helped with parts of your post, e.g. the team members who
obtained data or set up a server or developed a package you use.

### Tags

Each post can be tagged, as you can see in the metadata block at the top. I
suggest we consider the following tags as base tags:

* forecasting
* symptom surveys
* COVIDcast API
* COVIDcast map
* nowcasting
* data sources
* news (for announcements of new features, new models, etc.)
* packages (for R and Python packages)

---

*[Frida Gomam](https://example.com) is a member of the Delphi group and a Ph.D.
student in data wrangling at the University of Southern North Dakota at Hoople.*
