---
title: "COVID-19 Symptom Surveys through Facebook"
author: "Alex Reinhart and Ryan Tibshirani"
date: 2020-08-11
tags: ["symptom surveys"]
output:
  html_document:
    code_folding: hide
---

In collaboration with Facebook, along with a consortium of universities and
public health officials, we've been conducting surveys since April 2020 to
monitor the spread and impact of the COVID-19 pandemic in the United States. Our
survey, advertised through Facebook, gets about 74,000 responses **every day.**
Respondents provide information about COVID-related symptoms, contacts, risk
factors, and demographics, allowing us to examine county-level trends across the
US. We believe such a combination of *detail* and *scale* has never before been
available in a public health emergency. In this post, we'll explore some of our
initial survey findings, show how you can get access to the data, and highlight
some exciting new directions we're now pursuing.

<!--more-->

**Credits.** *Logan Brooks and Katie Mazaitis wrote the initial code for
processing the survey data and producing estimates. Taylor Arnold and Alex
Reinhart provided huge help with refactoring and improving this codebase. Logan
Brooks, Pratik Patil, Rob Tibshirani, and Ryan Tibshirani developed the
statistical methodology behind the estimates computed from the survey data. The
COVIDcast API, which serves these estimates and is updated daily, is a much
larger effort run by the Delphi group; and our entire engineering team is owed a
lot of credit here. Ryan Tibshirani came up with the idea of running the
surveys, and worked with Facebook to make this a reality. On the Facebook side,
Curtiss Cobb and Jonathan McKay played a big role.*

In collaboration with Facebook, along with a consortium of universities and
public health officials, we've been conducting surveys since April 2020 to
monitor the spread and impact of the COVID-19 pandemic in the United States. Our
survey is advertised through Facebook, but it's run on our own Qualtrics
platform (Facebook never sees any of the survey responses). This is an ongoing
operation and our survey gets about 74,000 responses every day. We believe that
such a combination of *detail* and *scale* has never before been available via
surveys in a public health emergency: we collect info on COVID-related symptoms,
contacts, risk factors, and demographics, enabling us to look at county-level
trends across the US. We make aggregated data publicly available daily through
our [COVIDcast
API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast.html), and we
make (fully de-identified) individual survey responses available to researchers
who agree to [data use
terms](https://dataforgood.fb.com/docs/covid-19-symptom-survey-request-for-data-access/).

In this blog post, we'll explore some of our initial findings with two
COVID-like illness indicators that are derived from our surveys. In terms of
what these surveys have to offer, this is really just the tip of the iceberg ...
and we'll highlight some other exciting directions that we're pursuing now.

## Short Background

Back in March 2020, we began discussions with Facebook about the idea of running
a survey, advertised through their site, to collect real-time, county-level
information on people experiencing COVID-like symptoms. The basic premise was
that we could use this information to construct a **leading indicator** of COVID
activity in the US. We certainly weren't the first group to think of running
COVID-19 symptom surveys: many groups before us had already deployed such
surveys (like CovidNearYou, COVID Symptom Study, etc.). But we sought to build a
survey that could (a) grow to be truly massive in size and (b) be sustained at
this size, something we felt was only possible with buy-in from a platform like
Facebook. Fortunately for us, they agreed!

We launched our survey on April 6, 2020 with Facebook. Each day since then,
Facebook directs a random sample of its users each day to our survey, hosted on
[Qualtrics](https://www.qualtrics.com/). As part of our agreement with Facebook,
we receive the data directly from Qualtrics, and Facebook never sees any of the
survey responses. As of this writing, our survey receives an average of over
74,000 responses per day; delivers adequate data (enough for us to create
meaningful estimates) for an average of nearly 1000 counties per week; and has
been taken over 9 million times so far.

We'll go into detail later about on our survey and our survey-based indicators,
but for now, here's a couple quick maps. On the left is a state-level heatmap
of the estimated percentage of people with COVID symptoms, based on our survey,
averaged over June 15, 2020 to July 15, 2020. On the right is a heatmap of the
number of daily new confirmed COVID-19 cases per 100,000 people, based on data
from [JHU CSSE](https://github.com/CSSEGISandData/COVID-19), averaged over the
same period. We can see some clear similarities in terms of the itensity
patterns, which is certainly a good sanity-check. (We did this all with our
[covidcast R package](https://cmu-delphi.github.io/covidcast/covidcastR/)---from
fetching the data from our API, to producing the heatmaps---and requires only
about 15 lines of code. You can click the "code" button to reveal it. We'll
cover our covidcast [R and Python
packages](https://cmu-delphi.github.io/delphi-epidata/api/covidcast_clients.html)
in a future blog post.)

```{r, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=5)
```

```{r, cache = TRUE, message = FALSE}
library(covidcast)
library(dplyr)
library(gridExtra)

# Fetch Facebook % CLI signal and JHU confirmed case incidence
start_date = "2020-06-15"
end_date = "2020-07-15"
df_fb = covidcast_signal("fb-survey", "smoothed_cli",
                         start_date, end_date, geo_type = "state")
df_in = covidcast_signal("jhu-csse", "confirmed_7dav_incidence_prop",
                         start_date, end_date, geo_type = "state")

# For each state, average the signals over all available times
df_fb_avg = df_fb %>% group_by(geo_value) %>% summarize(value = mean(value))
df_in_avg = df_in %>% group_by(geo_value) %>% summarize(value = mean(value))

# Set a bunch of fields so that the data frames know how to plot themselves
df_fb_avg$time_value = df_in_avg$time_value = end_date
df_fb_avg$issue = df_in_avg$issue = end_date
df_fb_avg$direction = df_in_avg$direction = 0
attributes(df_fb_avg)$geo_type = attributes(df_in_avg)$geo_type = "state"
class(df_fb_avg) = c("covidcast_signal", class(df_fb_avg))
class(df_in_avg) = c("covidcast_signal", class(df_in_avg))

# Plot choropleth maps, using the covidcast plotting functionality
subtitle = paste("Averaged over", start_date, "to", end_date)
p1 = plot(df_fb_avg,
          title = "% of people with COVID symptoms, based on Facebook surveys",
          range=c(0,1), choro_params = list(subtitle = subtitle))
p2 = plot(df_in_avg,
          title = "Daily new confirmed COVID-19 cases per 100,000 people",
          range=c(0,30), choro_params = list(subtitle = subtitle))
grid.arrange(p1, p2, nrow = 1)
```

## Why Run These Surveys?

Now let's unpack the main motivation behind our survey a bit: a person typically
experiences COVID-like symptoms before they seek medical care or a COVID-19
test, so data on how many people are self-reporting COVID-like symptoms in a
given county could be used to construct a leading indicator of COVID activity in
that county. And to be clear, it's not just the fact that we're looking at
symptoms that's important here, it's the *way* we're measuring them: since
self-reporting symptoms can be done from home, with no special equipment needed,
the data here wouldn't be subject to the same reporting delays as formal testing
metrics.

At this point it's also worth being very clear about what we're *not* able to
say with these surveys.

- Symptoms alone are not sufficient to diagnose coronavirus infections; of
course, COVID-like symptoms can be caused by other conditions, and many true
infections are asymptomatic. Therefore, even if we ignore the issues brought on
by self-reporting and survey sampling, we can't expect the estimates we produce
to be reflective of the true rate of COVID-19 (and they're not intended to).

- Our survey responses come from the population of Facebook users in the US,
which may a sizeable fraction of the US population, but certainly not all of it.
We'll also likely see survey bias because some people on Facebook are more
inclined to take surveys of this nature than others. We attempt to correct for
both of these biases using a [statistical reweighting
scheme](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/fb-survey.html#survey-weighting)),
but these corrections aren't perfect.

- Our symptom data is entirely self-reported (as opposed to, for example, being
reported by a medical professional). Some fraction of the responses could be
erroneous, either because a person misunderstood the question or chose to answer
incorrectly.

To summarize, our survey data can't be used to make absolute statements about
the true prevalence of coronavirus disease in the US; in fact, it shouldn't even
be regarded as a foolproof way of deriving unbiased estimates of the number of
people with COVID-19 *symptoms* in the US (mainly due to the self-reporting
aspect). Nevertheless, as we'll see evidence of below, *changes in self-reported
symptoms over time* can still be a meaningful reflection of the changes in
coronavirus infections over time. (And in the best case, it'll be a reflection
of changes to come some number of days into the future.)

## What's on the Survey?

Our survey has 4 sections and is about 35 questions long. The first section is
short and gathers info on a core set of symptoms used to define a condition
called **COVID-like illness** or **CLI**, which we define as *fever of at least
100 °F, along with cough, shortness of breath, or difficulty breathing*. This
mirrors the standard definition of influenza-like illness or ILI (defined as
fever of at least 100 °F, along with sore throat or cough), and is in line with
the working definition of CLI used by the CDC.

Using data from responses to the survey's first section, we can compute
estimates of two key quantities, in a given location, on a given day:

- % CLI: the percentage of people with COVID-like illness; and

- % CLI-in-community: the percentage of people who *know someone in their
local community* with COVID-like illness.

Details on how we compute the % CLI and % CLI-in-community estimates can be
found in our [COVIDcast signals
documentation](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/fb-survey.html).
The % CLI-in-community indicator has an interesting backstory: it was sort of a
"happy accident" of an experiment we tried with surveys we were running in
partnership with Google. (We'll cover our Google survey in a future blog post.)
The preliminary results from this experiment were exciting enough that we
decided to add a similar question to our Facebook survey about a week after its
initial launch.

The second section of the survey goes into more detail on symptoms, testing, and
medical-seeking behavior. The third section gathers info on contacts and risk
factors, and the fourth section gather info on demographics. These survey
questions, taken altogether, provide a wealth of information that we haven't
fully tapped into (but plan to soon---see the end of this blog post). To give
credit where it's due: it was Facebook who initially suggested we run a longer
survey with extensive questions on contacts, risk factors, and demographics. At
first, we wanted to run a short survey with only the questions on COVID-like
illness, and Facebook gathered ideas for more questions from several different
research groups in public health. They assured us that lengthening the survey
wouldn't hurt the the completion rate too much, and they were right: of the
people who complete the first section, over 85% go on to finish the full survey!

Our [survey documentation
site](https://cmu-delphi.github.io/delphi-epidata/symptom-survey/) includes more
details about the survey, including the [full text of every survey
version](https://cmu-delphi.github.io/delphi-epidata/symptom-survey/coding.html).
And yes, researchers can [request
access](https://dataforgood.fb.com/docs/covid-19-symptom-survey-request-for-data-access/)
to (fully de-identified) individual survey responses for research purposes.

## Some Interesting Examples

TODO data examples where the FB signal is a leading indicator. Start with
Miami-Dade? Then give the top 25 sort of thing?

Should we explain that the survey bias doesn't even have to be spatially
constant; for each county, say, the bias only has to be temporally constant or
even slowly drifting?

## Basic Correlation Analysis

TODO Compute correlations over time at multiple lags, plot them. Hopefully
maximized at some nonzero lag (lead).

Maybe do it on the scale of first differences, rather than values themselves?

## What's Next with the Surveys

TODO Alex can you write this part?

---

*[Alex Reinhart](https://www.refsmmat.com) manages Delphi's surveys, and is an
Assistant Teaching Professor in the Department of Statistics & Data Science.*

*[Ryan Tibshirani](http://stat.cmu.edu/~ryantibs) is a lead researcher in the
Delphi group, and an Associate Professor in the Department of Statistics & Data
Science and the Machine Learning Department at CMU. He is also an Amazon
Scholar.*
